---
title: "Module 7: Variable Creation and Classes"
format: revealjs
scrollable: true
smaller: true
---

## Learning Objectives

After module 7, you should be able to...

-   


# Adding/Removing Columns

## Adding new columns

You can add a new column, called `newcol` to `df`, using the `$` operator:
```{r}
df$newcol <- df$wt/2.2
head(df,3)
```

## Creating conditional variables

One frequently-used tool is creating variables with conditions. A general function for creating new variables based on existing variables is the Base R `ifelse()` function, which "returns a value depending on whether the element of test is `TRUE` or `FALSE`."

```{r, echo = FALSE, results = "asis"}
library(printr)
?ifelse
```


## `ifelse` example

```{r}
df$disp
```

Now with `ifelse`()
```{r}
#ifelse(test, yes, no)
ifelse(df$disp <= 200, "low", "high")
```


## Removing columns

You can remove a column by assigning to `NULL`:
```{r, eval = FALSE}
df$newcol <- NULL
```





## Summary

- Adding (or modifying) columns/variable to a data frame by using `$` 
- can remove a column in a few ways:  
    1. assigning a column to `NULL`  
    2. using `[]` with negative sign in front of column name(s)  
-   
-   


## Acknowledgements

These are the materials I looked through, modified, or extracted to complete this module's lecture.

-   ["Introduction to R for Public Health Researchers" Johns Hopkins University](https://jhudatascience.org/intro_to_r/)
-   [CRAN Project](https://cran.r-project.org/doc/manuals/R-lang.html#Indexing)
-   [CRAN Project](https://cran.r-project.org/web/packages/extraoperators/vignettes/logicals-vignette.html)

