---
title: "Module 10: Data Visualization"
format: 
  revealjs:
    scrollable: true
    smaller: true
---

## Learning Objectives

After module 10, you should be able to...

-		Create Base R plots

## Import data for this module

Let's read in our data (again) and take a quick look.

```{r echo=TRUE}
df <- read.csv(file = "data/serodata.csv") #relative path
head(x=df, n=3)
```

## Prep data

Create `age_group` three level factor variable
```{r echo=TRUE}
df$age_group <- ifelse(df$age <= 5, "young", 
                       ifelse(df$age<=10 & df$age>5, "middle", 
                              ifelse(df$age>10, "old", NA)))
df$age_group <- factor(df$age_group, levels=c("young", "middle", "old"))
```

Create `seropos` binary variable representing seropositivity if antibody concentrations are >10 mIUmL.
```{r echo=TRUE}
df$seropos <- ifelse(df$IgG_concentration<10, 0, 
										ifelse(df$IgG_concentration>=10, 1, NA))
```

## Base R data visualizattion functions

The Base R 'graphics' package has a ton of graphics options. 

```{r, echo = FALSE, results = "asis"}
library(printr)
?graphics
```

```{r echo=TRUE}
library(help = "graphics")
```


## Focus on a handful here today

`hist()` displays histogram of one variable
`plot()` displays x-y plot of two variables



## Summary

-		

## Acknowledgements

These are the materials I looked through, modified, or extracted to complete this module's lecture.

-   ["Introduction to R for Public Health Researchers" Johns Hopkins University](https://jhudatascience.org/intro_to_r/)
